Section Three

### 2020-10-19 new2020/sections/sec03.Rmd
some typing

To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE. This used the particle filter and network model described by @cite:Elliott_2020;, and the arrival time predictions described above. At the end of the day, we could compare the predicted with the actual arrival times. In addition, we compare to the GTFS-based predictions which are simply the scheduled arrival time plus the delay at the most recent stop.

@@

The overall values are shown in @cref:tab:overall_results;. The particle filter predictions are slightly more accurate than GTFS on average, and have about 20% error versus GTFS's 27%. However, only 77% of prediction intervals contain the actual arrival time, indicating that uncertainties are being under-predicted (the nominal coverage is 85%).

>>>>

The particle filter vehicle model and Kalman filter network model described by @citet:Elliott_2020; were combined with the particle filter prediction model described in \cref{particle-filter-etas} and applied to a full day of data from [[date]]. For each prediction of arrival time at each stop along all routes, we stored the mean along with the 5 and 90% quantiles, which could be compared to the bus's actual arrival time at the end of the day. We also computed the schedule-delay prediction using the real-time trip updates to compare our results to the status quo.

To assess and compare predictive performance, we use RMSE, MAE, MAPE, and PICP. For random variables $\{X_n: n=1,\ldots,M\}$ with estimates $\{\hat X_n: n=1,\ldots,M\}$,
\begin{equation*}
\begin{split}
\text{RMSE} &= \sqrt{\frac{1}{M} \sum_{n=1}^M \left(\hat X_n - X_n\right)^2},\\
\text{MAE} &= \frac{1}{M}\sum_{n=1}^M \left|\hat X_n - X_n\right|,\\
\text{MAPE} &= \frac{1}{M}\sum_{n=1}^M \frac{\left|\hat X_n - X_n\right|}{X_n}.
\end{split}
\end{equation*}
For prediction intervals $\{Q_n = (X_{na},X_{nb}):n=1,\ldots,M\}$, the indicator variable $I_n$ is 1 if $X\in Q_n$, and zero otherwise. Thus,
\[
\text{PICP} = \frac{1}{M} \sum_{i=1}^M I_n.
\]
[[ copy this definition from thesis ...]].

RMSE and MAE each evaluate the accuracy of the predictor, with MAE using absolute error is less influenced by extreme prediction errors. MAPE evaluates the proportional error of the predictor, which is appropriate for arrival time prediction as we expect errors to increase for stops farther away. Finally, PICP examines how well the model incorporates the underlying arrival time uncertainty of the model: we expect the PICP for an 85% prediction interval to be approximately 0.85. This value is not available for the schedule-delay predictor which only provides point estimates of arrival time.

<!-- To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE. This used the particle filter and network model described by @cite:Elliott_2020;, and the arrival time predictions described above. At the end of the day, we could compare the predicted with the actual arrival times. In addition, we compare to the GTFS-based predictions which are simply the scheduled arrival time plus the delay at the most recent stop.

@@

The overall values are shown in @cref:tab:overall_results;. The particle filter predictions are slightly more accurate than GTFS on average, and have about 20% error versus GTFS's 27%. However, only 77% of prediction intervals contain the actual arrival time, indicating that uncertainties are being under-predicted (the nominal coverage is 85%). -->
